Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-02 ms.tex
ahhhh

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases reside near the
nominal Bondi radius).  When heating is sufficiently weak, however, no stagnation radius may exist at all, i.e. the entire ISM of the galaxy is flowing towards the SMBH.  A goal of this work is to
assess under what conditions a stagnation radius exists, as this
question is intimately tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe some
analytic results, while in $\S\ref{sec:numerical}$ we describe our
numerical results.  In $\S\ref{sec:discussion}$ we discuss the
implications of our results, to topics including the nuclear X-ray luminosities of quiescent black holes, jetted TDEs, and the growth time of SMBHs.  In $\S\ref{sec:conclusions}$ we summarize our conclusions.  Table
\ref{table:definitions} provides the definitions of commonly used
variables.  Appendices \ref{app:rs} and \ref{app:be} provides useful
analytic results, while Appendix \ref{app:windheat} describes our
method for calculating stellar wind heating and mass input in greater
detail.

@@

We fix $\beta = 2$ (a typical value) because our results are generally
not sensitive to the properties of the gas flow on radial scales
$\gtrsim r_b$. The presence of the break radius $r_{b}$ is, however,
necessary to obtain a converged steady state for some regions of our
parameter space.  We consider solutions calculated for up to four
values of $\rb$: 50 pc, 100 pc, 200 pc, and 400 pc, motivated by the
range of break radii from the \citet{LauerFaber+:2007a}
sample.\footnote{For the core galaxies the break radius follows the
  scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with
  scatter of approximately one dex.  Most the cusp galaxies have $\rb$
  between 100 and 1000 pc, without a clear trend with $\Mbh$.}  In
practice we calculate solutions for a range of $\rb$ only in cases
when its precise value is expected to have a significant affect on the
solution, in particular for high SMBH masses and low heating rates,
for which the stagnation radius approaches large radii.

@@

Continuity of the entropy derivative at the stagnation radius where $v = 0$ requires that the temperature at this location be given by (eq.~[\ref{eq:first_law}])
\begin{align}
T|_{r_{s}}& = \frac{\gamma-1}{\gamma}\frac{\mu m_p
  \tilde{v}_{w}^{2}}{2k} \approx \frac{\gamma-1}{\gamma}\frac{\mu
  m_p v_w^2}{2k} \frac{3+8\Gamma}{3+8\Gamma-6\nu} \nonumber\\ 
 &\approx
 \begin{cases}
  5.0\times 10^6\ \, v_{500}^2 \,\,{\rm K} & \text{core} \\
  5.5\times 10^6\ \, v_{500}^2 \,\,{\rm K} & \text{cusp},
 \end{cases}
\label{eq:Tanalytic}
\end{align}
where we have assumed $\gamma=5/3$, $v_{500} \equiv v_{w}/(500$ km
s$^{-1}$), $v_w\gg\sigma_0$,  and $\densSlope \equiv -d{\rm ln}\rho/d{\rm ln\,r}|_{r_{\rm
    s}}$ is the density power-law slope at $r = r_{\rm s}$.
Empirically, we find that
\begin{align}
\densSlope \simeq \frac{1}{6} \left(4 \Gamma+3\right).
\label{eq:densSlope}
\end{align}

@@

The minimum heating rate for thermal stability (eq.~[\ref{eq:cooling2}]) corresponds to maximum thermally-stable accretion rate,
\begin{align}
\frac{\dot{M}_{\rm TI}}{\dot{M}_{\rm edd}} \simeq
  {\rm min}\begin{cases}
   5\times 10^{-4} \eta_{0.02}^{0.47} M_{\bullet,8}^{0.36} &, \text{core}\\
3\times 10^{-6}\eta_{0.02}M_{\bullet,8}^{-0.38}\\
   2\times 10^{-4} \eta_{0.02}^{0.59} M_{\bullet,8}^{0.28}    &, \text{cusp}\\
2.5\times 10^{-4}\eta_{0.02}
  \end{cases}
  \label{eq:Mdotmax}
\end{align}
where the first line for each case is derived using equation (\ref{eq:eddr_analytic}) to calculate
$\dot{M}(v_{\rm TI})$, whereas the second line corresponds to the low wind heating limit and is calculated for core galaxies using the empirical correlation $r_{b}/r_{\rm soi} \approx 7.6M_{\bullet,8}^{0.2}$ which follows from equation (\ref{eq:rsoi}) and the results of \citet{LauerFaber+:2007a}.

% For each of our galaxies Table \ref{tab:eta} provides the maximum value of $\eta$ such that $H/C >$1 at all radii.  Roughly speaking, this is the maximum $\eta$ for which it would be safe to ignore cooling.  If cooling is important, the CNM is likely to be thermally unstable, resulting in the formation of a cold medium that could substantially enhance the rate of accretion onto the SMBH.

%\begin{align}
 % \Lambda(T)\simeq
  %\begin{cases}
   % 2.3 \times 10^{-24} \left(T/10^6 \text{K}\right)^{0.5} $erg cm$^3 $s$^{-1}& \text{if } T \gsim 2 \times 10^7 \text{K} \\
   % 1.1 \times 10^{-22} \left(T/10^6 \text{K}\right)^{-0.7}  $erg cm$^3 $s$^{-1}& \text{if } T \lsim 2 \times 10^7 \text{K}     
  %\end{cases}
  %\label{eq:cooling}
%\end{align}

@@

Our numerical results, summarized in Table \ref{table:models}, allow us to study a range of CNM properties and to assess the validity of the analytic estimates from the previous section.  Cases for which $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|< 10$ {\bf BDM: correct?} at the stagnation radius are marked as `thermally unstable', as are cases for which the stagnation radius diverges and hence could not be resolved, resulting in their calculations being terminated.

Figure~\ref{fig:profiles} shows profiles of the density $\rho(r)$,
temperature $T(r)$, and radial velocity $|v(r)|$ (expressed in units of the sound speed), for the cusp
($\Gamma=0.8$) solutions within our grid.  As expected, the gas
density increases towards the SMBH $\rho\propto r^{-\densSlope}$ with
$\densSlope\simeq1$, i.e. shallower than the $-3/2$ power law for
Bondi accretion. This power law behavior does not extend
through all radii, however, as the gas density profile has a break coincident
with the location of the break in the stellar light profile ($\rb=100$
pc). The temperature profile is relatively flat at large radii, but
increases as $\propto 1/r^{k}$ interior to the sphere of influence,
where $k\lesssim 1$, somewhat shallower than expected for virialized gas within the black hole sphere of influence.
The inwardly directed velocity increases towards the hole with a profile that is somewhat steeper than the local free-fall velocity $v \propto v_{\rm ff}\propto r^{-1/2}$.  The flow near the stagnation radius is subsonic, but becomes supersonic at two critical points, one located at small radii, $r \lesssim 0.1r_{\rm s}$, and the other located at large radii, $r \gtrsim 10^{3}r_{\rm s}$.  In many cases the supersonic transition is instigated by the transition to a steeper stellar density profile exterior to the outer Nuker break radius, making our solutions near $r_{\rm s}$ also potentially sensitive to the break radius value.
% {\bf BDM: I made this up - in general we need
%   more explanation of what the plots are actually showing, with some
%   physical explanation.}
% AG -- things are not quite approaching Bondi on the inner grid. I
% don't have a good physical explanation yet...

@@

Our analytic estimates accurately reproduce the numerical results in the high heating limit $\zeta \gg 1$ ($v_w \gg \sigma_0$; $r_{\rm s} \lesssim r_{\rm soi}$).  However, for sufficiently low heating the stagnation radius diverges above the analytic estimate, approaching the stellar break radius $r_b \gg r_{\rm soi}$.  This divergence occurs approximately when $\zeta = \sqrt{v_w^{2}+\sigma_0^{2}}/\sigma_0 < \zeta_c \propto (r_b/r_{\rm soi})^{0.5(\Gamma-1)}$ (eq.~[\ref{eq:zetacrit}], Appendix \ref{app:be}), and hence occurs at a higher value of $\zeta$ in the case of core galaxies ($\Gamma = 0.8$) with large break radii.  This behavior explains the three core galaxies shown as green squares, which are vertically aligned at the same value of $\zeta$ close to where the analytic estimate diverges; these correspond to models with the same black hole mass ($\Mbh = 10^7 \Msun$) and heating rate ($v_w = 300$ km s$^{-1}$) but different values of the break radius: $r_{b} = $ 50, 100, and 400 pc (from top to bottom) corresponding to successively larger values of $\zeta_c$.

 Figure~\ref{fig:mdot_mass} shows the SMBH accretion rate for each of our numerical solutions in Eddington units for different values of $\vwO =$ 300, 600 and 1200 km s$^{-1}$, and for both core ($\Gamma$=0.1) and cusp galaxies ($\Gamma$=0.8).  Shown for comparison is our analytic estimate of $\dot{M}/\dot{M}_{\rm edd}$ from equation (\ref{eq:eddr_analytic}).  For high wind velocities ($v_{w} \gg \sigma$) the stagnation lies well inside the black hole sphere of influence and our analytic estimate provides a good fit to the numerical results.  However, for low wind velocities and/or high $M_{\bullet}$ (large $\sigma$), the numerical accretion rate considerably exceeds the analytic estimate as the stagnation radius diverges to large radii (Fig.~\ref{fig:stag}).  
%% AG: Describe conservation checks to additionally validate the results.

\begin{figure}
  \includegraphics[width=\columnwidth]{rs.eps}
  \caption{\label{fig:stag} Stagnation radius
    $r_{s}$ in units of the sphere of influence radius $r_{\rm soi}$
    (eq.~[\ref{eq:rsoi}]) for galaxies in our sample as a function of the stellar wind heating parameter $\zeta \equiv \sqrt{v_w^{2} + \sigma_0^{2}}/\sigma_0$.  Green, orange, and blue
    symbols correspond to different values of $v_{w} =$ 300, 600, and
    1200 km s$^{-1}$, respectively.  Squares correspond to cusp
    galaxies ($\Gamma = 0.8$), while triangles correspond to cores
    ($\Gamma = 0.1$). The black curves correspond to the analytic
    prediction from equation (\ref{eq:rs2main}), with solid and dashed
    curves calculated for parameters $\{\Gamma=0.8, \densSlope\simeq1\}$ and
    $\{\Gamma=0.1,\densSlope\simeq0.6\}$, respectively. }
\end{figure}

>>>>

In the presence of strong heating, one-dimensional steady-state flow
is characterized by an inflow-outflow structure, with a critical
radius known as the``stagnation radius" $\rs$ where the radial
velocity passes through zero.  Mass loss from stars interior to the
stagnation radius is accreted, while that outside $\rs$ is unbound in
an outflow from the nucleus.  The stagnation radius, rather than the
Bondi radius, thus controls the SMBH accretion rate (although we will
show that $\rs$ can in many cases resides near the nominal Bondi
radius).  When heating is sufficiently weak, however, the stagnation
radius may not exist at all, i.e. the entire ISM of the galaxy is flowing
towards the SMBH.  A goal of this work is to assess under what
conditions a stagnation radius exists, as this question is intimately
tied to that of cooling instability.
% Furthermore, in some cases we find that no stagnation
% radius exists at all; this formally implies that the entire of ISM of
% the galaxy is feeding the SMBH, potentially resulting in much higher
% accretion rate than would be predicted by simple Bondi accretion.

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxies used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe some
analytic results, while in $\S\ref{sec:numerical}$ we describe our
numerical results.  In $\S\ref{sec:discussion}$ we discuss the
implications of our results, to topics including the nuclear X-ray
luminosities of quiescent black holes, jetted TDEs, and the growth
time of SMBHs.  In $\S\ref{sec:conclusions}$ we summarize our
conclusions.  Table \ref{table:definitions} provides the definitions
of commonly used variables.  Appendices \ref{app:rs} and \ref{app:be}
provide useful analytic results for the stagnation radius and the
Bernoulli parameter of the flow, while Appendix
\ref{app:windheat} describes our method for calculating stellar wind
heating and mass input in greater detail.

@@

We fix $\beta = 2$ (a typical value) because our results are generally
not sensitive to the properties of the gas flow on radial scales
$\gtrsim r_b$. The presence of the break radius $r_{b}$ is, however,
necessary to obtain a converged steady state for some regions of our
parameter space.  We consider solutions calculated for up to four
values of $\rb$: 50 pc, 100 pc, 200 pc, and 400 pc, motivated by the
range of break radii from the \citet{LauerFaber+:2007a}
sample.\footnote{For the core galaxies the break radius follows the
  scaling relationship $\rb\sim 106 \, (\Mbheight)^{0.39}$ pc, with
  scatter of approximately one dex.  Most the cusp galaxies have $\rb$
  between 100 and 1000 pc, without a clear trend with $\Mbh$.}  In
practice we calculate solutions for a range of $\rb$ only in cases
when its precise value is expected to have a significant affect on the
solution, in particular for high SMBH masses and low heating rates,
for which the stagnation radius approaches large radii. {\bf AG
  describe unphysically large rb for the case of cores.}

@@

Continuity of the entropy derivative at the stagnation radius where $v = 0$ requires that the temperature at this location be given by (eq.~[\ref{eq:first_law}])
\begin{align}
T|_{r_{s}}& = \frac{\gamma-1}{\gamma}\frac{\mu m_p
  \tilde{v}_{w}^{2}}{2k} \approx \frac{\gamma-1}{\gamma}\frac{\mu
  m_p v_w^2}{2k} \frac{13+8\Gamma}{13+8\Gamma-6\nu} \nonumber\\ 
 &\approx
 \begin{cases}
  5.0\times 10^6\ \, v_{500}^2 \,\,{\rm K} & \text{core} \\
  5.5\times 10^6\ \, v_{500}^2 \,\,{\rm K} & \text{cusp},
 \end{cases}
\label{eq:Tanalytic}
\end{align}
where we have assumed $\gamma=5/3$, $v_{500} \equiv v_{w}/(500$ km
s$^{-1}$), $v_w\gg\sigma_0$,  and $\densSlope \equiv -d{\rm ln}\rho/d{\rm ln\,r}|_{r_{\rm
    s}}$ is the density power-law slope at $r = r_{\rm s}$.
Empirically, we find that
\begin{align}
\densSlope \simeq \frac{1}{6} \left(4 \Gamma+3\right).
\label{eq:densSlope}
\end{align}

@@

The minimum heating rate for thermal stability (eq.~[\ref{eq:cooling2}]) corresponds to maximum thermally-stable accretion rate,
\begin{align}
\frac{\dot{M}_{\rm TI}}{\dot{M}_{\rm edd}} \simeq
  {\rm min}\begin{cases}
   5\times 10^{-4} \eta_{0.02}^{0.47} M_{\bullet,8}^{0.36} &, \text{core}\\
3\times 10^{-6}\eta_{0.02}M_{\bullet,8}^{-0.38}\\
   2\times 10^{-4} \eta_{0.02}^{0.59} M_{\bullet,8}^{0.28}    &, \text{cusp}\\
2.5\times 10^{-4}\eta_{0.02}
  \end{cases}
  \label{eq:Mdotmax}
\end{align}
where the first line for each case is derived using equation
(\ref{eq:eddr_analytic}) to calculate $\dot{M}(v_{\rm TI})$, whereas
the second line corresponds to the low wind heating limit and is
calculated for core galaxies using the empirical correlation
$r_{b}/r_{\rm soi} \approx 7.6M_{\bullet,8}^{-0.2}$ which follows from
equation (\ref{eq:rsoi}) and the results of \citet{LauerFaber+:2007a}.

@@

Our numerical results, summarized in Table \ref{table:models}, allow
us to study a range of CNM properties and to assess the validity of
the analytic estimates from the previous section.

We include a few solutions with cooling (marked with a $\dagger$) and
conductivity (marked with a $\ddagger$) Table \ref{table:models} to
assess whether it is valid to neglect these effects. As expected
solutions for which $\dot{q}_{\rm heat}/\dot{q}_{\rm rad}\gg 1$ are
unaffected when radiative cooling is included (see the solution with
$\Mbh=10^{8} \Msun, v_w=600 {\rm km/s and} \gamma=0.8$. However, we
find that as we decrease the heating further, to cooling begins to
become more important (see $\Mbh=10^{8} \Msun, v_w=450 {\rm km/s and}
\gamma=0.8$). Overall, we find radiative cooling has negligible effect
on cusp solutions $v_w\geq 600 km/s$ in and core solutions with
$v_w=1200$ km/s {\bf AG for the purposes of description it may be
  simpler to just rerun all of our models with radiative cooling}.

We re-ran all cusps ($\Gamma=0.8$) with $v_w=300$ km/s and
all cores ($\Gamma=0.1$) with $v_w=600 $ km/s with radiative cooling
for three different values of the mass input parameter $\eta$--0.02,
0.2, and 0.6.  Note that $\eta=0.02$ is the minimum expected physical
mass input parameter. Solutions which were unstable for all values of
$\eta$ are marked with a $\mathbf{TI}$ {\bf solution that were
  unstable for only some choices of eta}.

Figure~\ref{fig:profiles} shows profiles of the density $\rho(r)$,
temperature $T(r)$, and radial velocity $|v(r)|$ (expressed in units
of the sound speed), for the cusp ($\Gamma=0.8$) solutions within our
grid.  As expected, the gas density increases towards the SMBH
$\rho\propto r^{-\densSlope}$ with $\densSlope\simeq1$, i.e. shallower
than the $-3/2$ power law for Bondi accretion. This power law behavior
does not extend through all radii, however, as the gas density profile
has a break coincident with the location of the break in the stellar
light profile ($\rb=100$ pc). The temperature profile is relatively
flat at large radii, but increases as $\propto 1/r^{k}$ interior to
the sphere of influence, where $k\lesssim 1$, somewhat shallower than
expected for virialized gas within the black hole sphere of influence.
The inwardly directed velocity increases towards the hole with a
profile that is somewhat steeper than the local free-fall velocity $v
\propto v_{\rm ff}\propto r^{-1/2}$.  The flow near the stagnation
radius is subsonic, but becomes supersonic at two critical points, one
located at small radii, $r \lesssim 0.1r_{\rm s}$, and the other
located at large radii, $r \gtrsim 10^{3}r_{\rm s}$.  In many cases
the supersonic transition is instigated by the transition to a steeper
stellar density profile exterior to the outer Nuker break radius,
making our solutions near $r_{\rm s}$ also potentially sensitive to
the break radius value.

@@

Our analytic estimates accurately reproduce the numerical results in
the high heating limit $\zeta \gg 1$ ($v_w \gg \sigma_0$; $r_{\rm s}
\lesssim r_{\rm soi}$).  However, for sufficiently low heating the
stagnation radius diverges above the analytic estimate, approaching
the stellar break radius $r_b \gg r_{\rm soi}$.  This divergence
occurs approximately when $\zeta =
\sqrt{v_w^{2}+\sigma_0^{2}}/\sigma_0 < \zeta_c \propto (r_b/r_{\rm
  soi})^{0.5(\Gamma-1)}$ (eq.~[\ref{eq:zetacrit}], Appendix
\ref{app:be}), and hence occurs at a higher value of $\zeta$ in the
case of core galaxies ($\Gamma = 0.8$) with large break radii.

As $\zeta$ approached $\zeta_c$, the results the location of the
stagnation radius will depend on the location of the break radius
$\rb$.  This explains the behavior of the two core galaxies, shown as
orange triangles vertically aligned at the same value of
$\zeta$. These correspond to models with the same black hole mass
($\Mbh = 10^8 \Msun$) and heating rate ($v_w = 600$ km s$^{-1}$) but
different values of the break radius: $r_{b} =$ 25, 50 pc (from top to
bottom). {\bf AG still needs work}

Figure~\ref{fig:mdot_mass} shows the SMBH accretion rate for each of
our numerical solutions in Eddington units for different values of
$\vwO =$ 300, 600 and 1200 km s$^{-1}$, and for both core
($\Gamma$=0.1) and cusp galaxies ($\Gamma$=0.8).  Shown for comparison
is our analytic estimate of $\dot{M}/\dot{M}_{\rm edd}$ from equation
(\ref{eq:eddr_analytic}).  For high wind velocities ($v_{w} \gg
\sigma$) the stagnation lies well inside the black hole sphere of
influence and our analytic estimate provides a good fit to the
numerical results.  However, for low wind velocities and/or high
$M_{\bullet}$ (large $\sigma$), the numerical accretion rate
considerably exceeds the analytic estimate as the stagnation radius
diverges to large radii (Fig.~\ref{fig:stag}).

\begin{figure}
  \includegraphics[width=\columnwidth]{rs.eps}
  \caption{\label{fig:stag} Stagnation radius $r_{s}$ in units of the
    sphere of influence radius $r_{\rm soi}$ (eq.~[\ref{eq:rsoi}]) for
    galaxies in our sample as a function of the stellar wind heating
    parameter $\zeta \equiv \sqrt{v_w^{2} + \sigma_0^{2}}/\sigma_0$.
    Green, orange, and blue symbols correspond to different values of
    $v_{w} =$ 300, 600, and 1200 km s$^{-1}$, respectively.  Squares
    correspond to cusp galaxies ($\Gamma = 0.8$), while triangles
    correspond to cores ($\Gamma = 0.1$). Green crosses correspond to
    cusp solutions which would be thermally unstable. The black curves correspond
    to the analytic prediction from equation (\ref{eq:rs2main}), with
    solid and dashed curves calculated for parameters $\{\Gamma=0.8,
    \densSlope\simeq1\}$ and $\{\Gamma=0.1,\densSlope\simeq0.6\}$,
    respectively. }
\end{figure}
